# Birds Data

This data set is on bird assemblages in a montane forest in the Velka Fatra Mountains (Slovak Republic).

```{r, message=FALSE}
library("readr")
library("tidyr")
library("dplyr")
library("tibble")
```

We can load the bird data and process it into two data sets

1. the species abundances, and
2. the associated environmental variables.


```{r}
birds_url <- "https://bit.ly/maed-birds"

birds <- read_csv(birds_url, na = "", skip = 1)

birds_spp <- birds %>%
    mutate(across(AegiCaud:OriOri, ~ replace_na(.x, 0))) %>%
    column_to_rownames("...1")
birds_spp <- birds_spp[, 1:36]

birds_env <- birds %>%
    select(c("...1", Altit:Expos)) %>%
    mutate(Rocks = factor(Rocks), Expos = factor(Expos)) %>%
    column_to_rownames("...1")
```

The aim of the analysis is to find a good ordination of the species data (choose your ordination method, transformation, dissimilarity, etc according to what features of the species data you want to focus on). Then, having found a good ordination, use the environmental data to try to explain the ordination axes or mapping you have created.

The variables are:

* Altitude
* Forest cover
* Forest density
* Amount of broad leaf trees
* E2 layer (scrub layer cover)
* E2 layer conifers
* E1 layer (herb layer cover)
* E1 height
* Slope
* Presence of rocks
* Exposure


# How the Canoco developers would analyse these data

They treat the data as "compositional" (despite them being the average number of breeding pairs averaged over 4 visits to each sampling location).

```{r}
birds_pca <- rda(birds_spp)

birds_pca

summary(eigenvals(birds_pca))

plot(birds_pca)

ev_pca <- envfit(birds_pca ~ Altit, data = birds_env)
ev_pca
plot(ev_pca, add = TRUE)

ordisurf(birds_pca ~ Altit, data = birds_env, add = TRUE)
```

# How I might have approached this

```{r}
birds_ca <- cca(birds_spp)

birds_ca

summary(eigenvals(birds_ca))

plot(birds_ca)

ev_ca <- envfit(birds_ca ~ Altit, data = birds_env)
ev_ca
plot(ev_ca, add = TRUE)

ordisurf(birds_ca ~ Altit, data = birds_env, add = TRUE)
```

or

```{r}
birds_h <- rda(decostand(birds_spp, method = "hellinger"))

birds_h

summary(eigenvals(birds_h))

plot(birds_h)

ev_h <- envfit(birds_h ~ Altit, data = birds_env)
ev_h
plot(ev_h, add = TRUE)

ordisurf(birds_h ~ Altit, data = birds_env, add = TRUE)
```

or


```{r}
birds_nmds <- metaMDS(decostand(birds_spp, method = "hellinger"),
    dist = "euclidean",
    autotransform = FALSE, expand = FALSE)

birds_nmds

stressplot(birds_nmds)

ordipointlabel(birds_nmds)
ev_nmds <- envfit(birds_nmds ~ Altit, data = birds_env)
ev_nmds

plot(ev_nmds, add = TRUE)

ordisurf(birds_nmds ~ Altit, data = birds_env, add = TRUE)
```

But does it make any difference?

```{r}
layout(matrix(1:4, ncol = 2, byrow = TRUE))
plot(birds_pca, main = "PCA")
plot(ev_pca, add = TRUE)
ordisurf(birds_pca ~ Altit, data = birds_env, add = TRUE,
    col = "forestgreen")
plot(birds_ca, main = "CA")
plot(ev_ca, add = TRUE)
ordisurf(birds_ca ~ Altit, data = birds_env, add = TRUE,
    col = "forestgreen")
plot(birds_h, main = "PCA (Hellinger)")
plot(ev_h, add = TRUE)
ordisurf(birds_h ~ Altit, data = birds_env, add = TRUE,
    col = "forestgreen")
ordipointlabel(birds_nmds, main = "NMDS")
plot(ev_nmds, add = TRUE)
ordisurf(birds_nmds ~ Altit, data = birds_env, add = TRUE,
    col = "forestgreen")
layout(1)
```