---
title: Permutation tests answers
author: Gavin Simpson
date: May 30, 2024
output: html_document
editor_options: 
  chunk_output_type: console
---

# Abstract

Below, you'll find several examples of experimental designs or other multivariate data sets that we wish to test using constrained ordination methods and restricted permutation (Monte Carlo) tests.

# Ectomycorrhizal fungi

In this example we will study the effect of sod-cutting and sod-addition on the number of sporocarps of ectomycorrhizal fungi in Pine stands differing in age and soil type using data from an experiment reported in Baar & Ter Braak (1996). The experimental layout is as follows. Six stands (St1&ndash;St6) with Scots pine of different age and soil type were selected. Within each stand, plots were laid out, with a size of 15 m x 15 m. Treatments applied to stands were:

* `S` &mdash; sod cutting, in which the litter and humus layers and the herbaceous vegetation were removed,

* `A` &mdash; sod addition, in which removed litter and humus were added onto the existing litter and humus layers, and

* `C` &mdash; control, in which the litter and humus layers were left unchanged.

Treatment `A` (sod addition) was not done in the oldest two stands (`St5` and `St6`) as these were the oldest stands and already had thick layers of humus and litter. The treatments were replicated 4 times each, hence in the younger stands we have 12 plots (4 replicates by 3 treatments), while in the oldest stands we have 8 plots each (4 replicates by 2 treatments). Within stands, treatments were randomized.

```{r}
library("vegan")
library("permute")
library("readr")
library("tidyr")
library("dplyr")
library("tibble")
library("forcats")
library("stringr")

fungi <- read_csv(url("https://bit.ly/fungi-spp"), skip = 1L,
    col_types = "-cddddddddddddddddddddddddddddddddddd") %>%
    mutate(across(Amangemm:Xerosubt, ~ replace_na(.x, 0))) %>%
        select(-(Ncarp:Nspec)) %>%
        rename(label = "...2")

design <- read_csv(url("https://bit.ly/fungi-design"), skip = 1L,
    col_types = "-cccdc") %>%
    rename(label = "...2", Treatment = `LH-treat`) %>%
        mutate(Treatment = fct_relevel(Treatment, "C", after = 0L),
            Treatment = fct_recode(Treatment, control = "C",
                sod_cutting = "S", sod_addition = "A"),
            Stand = factor(Stand),
            SoilType = factor(SoilType))
```


## Testing the effects of whole-plot variables

### Testing the effects of stand age and soil type

Using restricted permutations for split-plot designs, we can test the effects of variables that vary between whole-plots (but which are constant at the split-lot level). However, this is only possible if we make the design balanced by

1. deleting all plots from stands `St5` and `St6`, or
2. deleting all the `sod_addition` plots.

For this particular hypothesis we are testing the effect of a variable that variables at the whole-plot level (the stand level), it is better to keep as many stands as possible, so the `sod_addition` plots will be deleted.

```{r delete-sod-addition-plots}
fungi_bal <- fungi %>%
    left_join(design) %>%
    filter(Treatment != "sod_addition") %>%
    select(label:"Xerosubt") %>%
    mutate(across(-label, ~ log1p(.x))) %>%
    column_to_rownames("label")

design_bal <- design %>%
    filter(Treatment != "sod_addition") %>%
    column_to_rownames("label")
```


* Fit the RDA using the two variables that vary at the stand level.

```{r whole-plot-model}
m1 <- rda(fungi_bal ~ Age + SoilType, data = design_bal)
m1

RsquareAdj(m1)
```

* To test variables at this whole-plot level we need to permute the whole plot while leaving the split-plots (samples within the whole-plots) unpermuted. Created the necessary design using `how()` to test the effects of the whole plot variables that only vary at the stand level (whole-plots).

```{r whole-plot-perm-design}
whole_h <- with(design_bal,
    how(plots = Plots(Stand, type = "free"),
        within = Within(type = "none", constant = FALSE),
        nperm = 999))
```

```{r whole-plot-test}
set.seed(42)
anova(m1, permutations = whole_h)
```

### Testing the effects of stand age adjusting for soil type

We can test the effect of `Age` while controlling (or adjusting) for the effect of `SoilType` by partialling out the effect of `SoilType` using `Condition()`.

* Fit an RDA to estimate the effect of `Age` while adjusting for the effect of `SoilType`

```{r whole-plot-age-adjusted-for-soil-type}
m2 <- rda(fungi_bal ~ Age + Condition(SoilType), data = design_bal)

m2
```

* Use an appropriate restricted permutation design to test the effects of `Age` adjusted for `SoilType`

```{r whole-plot-adjusted-test}
set.seed(42)
anova(m2, permutations = whole_h)
```

## Testing the effect of split-plot variables

The treatment varies at the split-plot (within) level. We now want to test the effects of the treatment on the abundance of the fungi. This is best done by randomising samples within blocks (stands), while keeping the blocks fixed. To do this correctly, we also need to include the block variable as a covariate to be partialled out using `Condition()`.

We can use all the data for this test, so we prepare the full data set for analysis

```{r prep-fungi-all}
fungi_all <- fungi %>%
    mutate(across(-label, ~ log1p(.x))) %>%
    column_to_rownames("label")

design_all <- design %>%
    column_to_rownames("label")
```

* Create an RDA to estimate the effect of the treatment while adjusting for the dependence structure in the data.


```{r fungi-split-plot}
m3 <- rda(fungi_all ~ Treatment + Condition(Stand), data = design_all)

m3
```

* Design an appropriate restricted permutation design to test the estimated treatment effect.

```{r split-plot-perm-design}
split_h <- with(design_all,
    how(blocks = Stand,
        within = Within(type = "free"),
        nperm = 999))
```

```{r split-plot-adjusted-test}
set.seed(42)
anova(m3, permutations = split_h)
```

Additionally, we might be interested in testing the effect of the treatment differs among stands of different age, adjusting for a possible treatment by soil type interaction.

* Fit an RDA to the data to test for different treatment effects among stands of different ages

```{r}
m4 <- rda(fungi_all ~ Treatment:Age + 
            Condition(Stand + Treatment + Treatment * SoilType),
          data = design_all)


m4

anova(m4, permutations = split_h)

m4 <- rda(fungi_all ~ Age + Treatment + Treatment:Age + 
            Condition(Stand + SoilType + Treatment:SoilType),
          data = design_all)


anova(m4, permutations = split_h, by = "margin")
```

* Using an appropriate permutation design, test the estimated treatment effects among stands of different ages

```{r}
set.seed(42)
anova(m4, permutations = split_h)
```


# Testing a spatial design

In this example you'll look at the effect of pH on plant species in a wetland. The samples were laid out in grid of 20 samples arranged in **five** rows and **four** columns over the wetland.

The response data are abundances of 100 plant species at each of the 20 plots. We also have some associated data to go with the species abundances. The plots were sampled in 1977 and 1988, and pH and water depth were recorded at each location in the grid for each year.

For this example we'll look only at the effect of pH on wetland plant abundance in 1988, so we will remove the older samples from the data before proceeding. We also need to ensure that the data are arranged in the data set in grid-column order. The code below loads the data, filters to retain the 1988 samples only, and then orders the data in grid-column order.

```{r}

grid_veg <- read_csv(url("https://bit.ly/grid-veg"),
    skip = 1L) %>%
    mutate(across(ACHILMIL:`POPUL-SP`, ~ replace_na(.x, 0))) %>%
        select(-("...1")) %>%
        rename(label = "...2")



grid <- read_csv(url("https://bit.ly/grid-predictors"), skip = 1L,
    col_types = "-cccdddd") %>%
    rename(label = "...2") %>%
    mutate(Column = factor(str_sub(Plot, 2L, 2L)),
           Plot = factor(Plot),
           Row = factor(Row)) %>%
    relocate(Column, .after = Row) %>%
    arrange(Year, Column, Row)

# grid 1988 data
g88_veg <- grid %>%
    left_join(grid_veg) %>%
    filter(Year == 1988) %>%
    select(c(label, ACHILMIL:`POPUL-SP`)) %>%
    column_to_rownames("label")

g88 <- grid %>%
    filter(Year == 1988) %>%
    column_to_rownames("label")

g88_h <- how(within = Within(type = "grid", nrow = 5, ncol = 4,
    mirror = TRUE))

# test
perm <- shuffle(nrow(g88), control = g88_h)
tmp <- matrix(0, nrow = 5, ncol = 4)
tmp[] <- as.character(g88$Plot)
tmp

tmp[] <- as.character(g88$Plot)[perm]
tmp # works!!

g1 <- rda(g88_veg ~ pH88, data = g88)
g1

set.seed(88)
anova(g1, permutations = g88_h)

# or
set.seed(42)
perms <- shuffleSet(nrow(g88), 499, control = g88_h, check = FALSE)
anova(g1, permutations = perms)
```

This leaves us with 20 samples arranged in a grid.

We want to look at how the species abundances vary with `pH`, while controlling for potential spatial autocorrelation across the wetland. Fit an appropriate constrained ordination to estimate the effect of `pH` on species abundances, and test the estimated `pH` effect using an appropriate restricted permutation test.

# Ohraz example

The code below loads the Ohraz data and the experimental design data using slightly more modern code than was used in the slides.
Repeated observations of composition from an experiment

* Factorial design (3 replicates)
* Treatments: fertilisation, mowing, *Molinia* removal

Results in 2 x 2 x 2 = 8 combinations in three replicates leading to 24 2m x 2m plots.

The data a repeated observations that include the baseline (before treatment), the interaction of treatment and time is of particular interest.

Test 1 of the hypotheses

> There are *no* directional changes in species composition in time that are common to all treatments or specific treatments

```{r ohraz}
library("vegan")
library("readr")
library("dplyr")

spp <- read_csv(url("https://bit.ly/ohraz-spp")) %>%
    rename(label = "...1") %>%
    janitor::clean_names()

molinia <- spp %>%
    select(label:molicaer)

spp <- spp %>%
    select(-molicaer) %>%
    tibble::column_to_rownames("label")

env <- read_csv(url("https://bit.ly/ohraz-env")) %>%
    rename(label = "...1") %>%
    mutate(across(c(mowing:removal, plotid), ~ factor(.x))) %>%
    tibble::column_to_rownames("label")
```

When we test for the interactions mentioned above, we'll use `plotid` as a covariate (conditioning variable) when setting up the model.

We fit an RDA
```{r}
a1 <- rda(spp ~ year + year:mowing + year:fertilizer + year:removal + Condition(plotid),
    data = env)
a1
```

To set up the appropriate permutation design we need to restrict permutations to be *within* the 2m x 2m plots. We can do this using the `blocks` argument, or via the `plots` argument. As some of the other hypotheses we might address with these data requires permuting under a split plot design, we will use the `plots` formulation to test H1
```{r}
h1 <- with(env,
    how(nperm = 999,
        within = Within(type = "free"),
        plots = Plots(strata = plotid, type = "none")))
```

```{r}
set.seed(42)
anova(a1, permutations = h1, model = "reduced")
```

Hypothesis 2 is

> The temporal trend in species is independent of the treatment. I.e. the individuals treatments do not differ in their temporal dynamics

To address this, we need to remove the average effect of `year` and focus on the `year` effects *within* each treatment.

```{r}
a2 <- rda(spp ~ year:mowing + year:fertilizer + year:removal + Condition(year + plotid),
    data = env)
a2
```

For this hypothesis, under the null hypothesis (no effect) the temporal changes in the individual plots are independent of the treatments and are thus exchangeable. However, we should keep the samples for each plot together (i.e. not permute *within* each plot). This design requires the use of `plots`; `blocks` are never permuted, so we can't shuffle entire blocks of samples using that mechanism. To create the correct design-based permutation test then we use
```{r}
h2 <- with(env,
    how(nperm = 999,
        within = Within(type = "none"),
        plots = Plots(strata = plotid, type = "free")))
```

Now we can use `anova()` to test hypothesis 2
```{r}
set.seed(42)
anova(a2, permutations = h2, model = "reduced")
```

The remaining 3 hypotheses are

> Fertilisation has no effect on the temporal changes in species composition

> Mowing has no effect on the temporal changes in species composition

> Removal has no effect on the temporal changes in species composition

To answer each of these hypotheses, would require us to modify `a2` such that all but one of the `year` - treatment interactions is removed from the model and added to the `Condition()`, e.g.

```{r}
a3 <- rda(spp ~ year:mowing + Condition(year + year:fertilizer + year:removal + plotid),
    data = env)
a3
```

We can avoid having to do this by hand for each hypothesis by recognising that the `by = "margin"` option of `anova()` is going to do this for us. We can replicate the results generated by Canoco presented in the text book using:

```{r}
set.seed(42)
anova(a2, by = "margin", permutations = h2, model = "reduced", parallel = 4)
```

# Hierarchical effects of crayfish

The code below loads the `crayfish` data and the scale predictor variables (`design`) using slightly more modern code than was used in the slides.

```{r crayfish}
crayfish <- read_csv(url("https://bit.ly/crayfish-spp")) %>%
    rename(label = "...1") %>%
    janitor::clean_names() %>%
    filter(!is.na(label)) %>%
    column_to_rownames("label")

design <- read_csv(url("https://bit.ly/crayfish-design")) %>%
    rename(label = "...1") %>%
    filter(!is.na(label)) %>%
    mutate(across(Watershed:`Run Nested`, ~ factor(.x))) %>%
    column_to_rownames("label")
```

Repeat the exercise from the slides, testing the different levels of the hierarchical scales of variation in crayfish species abundance.

A number of samples have 0 crayfish, which excludes unimodal methods, hence we will use RDA for this analysis.

## Crayfish &mdash; Watershed scale

We first start at the watershed scale

```{r crayfish-watershed}
m.ws <- rda(crayfish ~ Watershed, data = design)
m.ws

summary(eigenvals(m.ws, constrained = TRUE))

set.seed(1)
ctrl <- how(nperm = 499, within = Within(type = "none"),
            plots = with(design, Plots(strata = Stream, type = "free")))
anova(m.ws, permutations = ctrl)
```

## Crayfish &mdash; Stream scale

```{r crayfish-stream}
m.str <- rda(crayfish ~ Stream + Condition(Watershed), data = design)
m.str

summary(eigenvals(m.str, constrained = TRUE))

set.seed(1)
ctrl <- how(nperm = 499, within = Within(type = "none"),
    plots = with(design, Plots(strata = Reach, type = "free")),
    blocks = with(design, Watershed))

anova(m.str, permutations = ctrl)
```

## Crayfish &mdash; Reach scale

```{r crayfish-reach}
m.re <- rda(crayfish ~ Reach + Condition(Stream), data = design)
m.re

summary(eigenvals(m.re, constrained = TRUE))

set.seed(1)
ctrl <- how(nperm = 499, within = Within(type = "none"),
            plots = with(design, Plots(strata = Run, type = "free")),
            blocks = with(design, Stream))
anova(m.re, permutations = ctrl)
```

## Crayfish &mdash; Run scale

```{r crayfish-run}
m.run <- rda(crayfish ~ Run + Condition(Reach), data = design)
m.run

summary(eigenvals(m.run, constrained = TRUE))

set.seed(1)
ctrl <- how(nperm = 499, within = Within(type = "free"),
            blocks = with(design, Reach))
anova(m.run, permutations = ctrl)
```

# Effect of ploughing time on weeds

Post (1986) carried out an experiment to investigate the effect of time of ploughing on the subsequent weed composition in summer barley fields. Three different ploughin times were used. There are 13 weed species observed in 12 plots in a randomized block experiment of four complete blocks of three plots each.

Traditionally, a multivariate ANOVA could have been used to analyses data like this. However, as there are more species (response variables; 13) than there are samples (12), MANOVA cannot be used. Instead, RDA is a useful option.

```{r plough-load}
plough <- read_csv(url("https://bit.ly/plough-spp"), skip = 1) %>%
    rename(label = "...1") %>%
    select(-`...2`) %>%
    janitor::clean_names() %>%
    filter(!is.na(label)) %>%
    tibble::column_to_rownames("label")

plough_design <- read_csv(url("https://bit.ly/plough-design"), skip = 1) %>%
    rename(label = "...1", ploughing_time = "Treatmnt") %>%
    select(-`...2`) %>%
    janitor::clean_names() %>%
    filter(!is.na(label)) %>%
    tibble::column_to_rownames("label")
```

We are interested in the effect of ploughing time and wish to remove any possible effect of the blocks. We achieve this by using `block` as a covariate (in `Condition()`). As the data are counts, a log-transformation is suggested (because we can't do a multivariate Poisson GLM to analyses these data.) Some zero abundances are recorded as `NA` values, so we must replace those with 0s before we proceed.

```{r}
plough <- plough %>%
    mutate(across(che_alb:ape_spi, .fns = tidyr::replace_na, replace = 0))
log_plough <- log1p(plough)
```

The model is specified as

```{r}
ord <- rda(log_plough ~ ploughing_time + Condition(block), data = plough_design)
ord
```

We see that almost half the variation in the species composition across the samples is explained by block-to-block differences. Ploughing time explains ~22% of the total variation in log species composition, and ~41% (2.5286 / (11.3151 - 5.2128)) of the variation after removing the block-to-block differences.

A correct permutation test for these data should keep samples contained at the block level and only permute within, not between, blocks.

```{r}
h <- with(plough_design,
    how(blocks = block, nperm = 999))
```

Using this design we can test the effect of ploughing time

```{r}
anova(ord, permutations = h)
```

Notice how we get a warning/message from *permute*

> Set of permutations < 'minperm'. Generating entire set.

This means that there are not a large number of permutations of these data owing to the block restriction. In such circumstances, if we randomly generated the permutations we would get many repeated permutations. Instead, it is better to generate the entire set of permutations and do an exact test. Hence, even though we asked for 999 permutations, the results indicate that 1295 permutations were used. This corresponds to the number of possible permutations of the data
```{r}
# this includes the observed ordering
numPerms(log_plough, control = h)
```

Another option for `margin` that is new in recent versions of *vegan*, is `"onedf"`. This will perform tests of one degree of freedom contrasts. For these data it means testing the *difference* between ploughing time 1 and 2, and between ploughing time 1 and 3

```{r}
anova(ord, permutations = h, by = "onedf")
```

This isn't so useful here, but if you had a control level and it was coded as the reference level of the factor, this would give you tests of control versus treatment level contrasts.

# A multivariate experiment with fixed factors

This example uses a data set of Swedish pine forest undergrowth assemblages that were experimentally manipulated through the addition of nitrogen `N` and phosphorus `P`. There are 32 samples in 4 complete, randomized bloacks of 8 plots each. In each block, all combinations of 4 levels of nitrogen `N` (applied as ammonium nitrate) and 2 levels of phosphorus `P` (applied as compound PK fertilizer) were applied.

```{r e40-load}
library("tidyr")
library("dplyr")
library("readr")

e40_spp <- read_csv(url("https://bit.ly/e40-spp"), skip = 1L) %>%
    rename(label = "...1") %>%
    select(-`...2`) %>%
    mutate(across(-label, ~ replace_na(.x, 0)),
           across(-label, ~ log1p(.x))) %>%
    tibble::column_to_rownames("label")

e40_design <- read_csv(url("https://bit.ly/e40-design"), skip = 1L) %>%
    rename(label = "...1") %>%
    select(-`...2`) %>%
    mutate(across(-label, ~ factor(.x))) %>%
    tibble::column_to_rownames("label")
```

The levels of `N` are

```{r}
with(e40_design, levels(N))
```

where `N0` means no N addition

The levels of `P` are

```{r}
with(e40_design, levels(P))
```

where `P0` means no P addition.

The variable `Block` codes for the 4 blocks of 8 plots.

In 1987 the undergrowth of the plots was surveyed. The percentage cover of plant species (including tree saplings, bryophytes, and lichens) was estimated and recorded on a ten-point scale. There are 103 species recorded. The ten-point scale values were re-expressed as the mid-points of the ten-point percentage scale in `e40_spp`. The species data are `log(x + 1)` transformed.

The aim of the analysis is to test the effects of

1. the N addition effect,
2. the P addition effect,
3. the combined effects of N and P additions
4. the magnitude of the interaction effect

using a combination of models and appropriate permutation tests.

## Testing the main effects

## Testing the N addition effect

```{r e40-n-addition-effect}
m_n <- rda(e40_spp ~ N + Condition(Block + P), data = e40_design)
m_n

h_blk <- with(e40_design, how(blocks = Block, nperm = 999))

set.seed(245)
anova(m_n, permutations = h_blk)
```

## Testing the P addition effect

```{r e40-p-addition-effect}
m_p <- rda(e40_spp ~ P + Condition(Block + N), data = e40_design)
m_p

set.seed(48)
anova(m_p, permutations = h_blk)
```

## Using `by = "margin"`

```{r}
# or
m_margin <- rda(e40_spp ~ N + P + Condition(Block), data = e40_design)
m_margin

set.seed(23)
anova(m_margin, permutations = h_blk, by = "margin")
```

## Omnibus effect of N and P

```{r}
m_np <- rda(e40_spp ~ N + P + Condition(Block), data = e40_design)
m_np

set.seed(23)
anova(m_np, permutations = h_blk)
```

## Testing the interaction effect - model based permutation

```{r e40-interaction}
m_int <- rda(e40_spp ~ N:P + Condition(Block + N + P), data = e40_design)
m_int

h_blk <- with(e40_design, how(blocks = Block, nperm = 999))

set.seed(42)
anova(m_int, permutations = h_blk)
```

```{r e40-interaction-v2}
m_int2 <- rda(e40_spp ~ N + P + N:P + Condition(Block), data = e40_design)
m_int2

set.seed(42)
anova(m_int2, permutations = h_blk, by = "margin")
```

## Testing the interaction effect - design based permutation

In general, design-based tests of interaction terms are not possible. However, for simple two-way designs such as the one we have here, it is possible to test for an interaction with a design-based test. For this to work we need to think of the data are being split into 4 blocks each with

* 4 columns, 1 per level of `N`
* 2 rows, 1 per level of `P`

And the ordering of the levels of `N` and `P` must be the same within each block of samples. If the data are arranged precisely like this, then we can permute the rows and the columns of these blocks of data to test the interaction effect.

Arrange the data in the correct order

```{r}
e40 <- e40_spp %>%
    bind_cols(e40_design) %>%
    group_by(Block) %>%
    arrange(N, P, .by_group = TRUE) %>%
    ungroup()

e40_spp2 <- e40 %>%
    select(agroscap:veronoff)

e40_design2 <- e40 %>%
    select(N:Block)

ord <- rda(e40_spp2 ~ N:P + Condition(N + P + Block), data = e40_design2)

h_int <- with(e40_design2,
    how(nperm = 999, plots = Plots(strata = N, type = "free"),
        blocks = Block,
        within = Within(type = "free")))

anova(ord, permutations = h_int)
```