--- 
title: "Analysing a randomised complete block design with vegan"
author: Gavin Simpson
date: May 30, 2024
output: html_document
---

```{r setup-options, echo = FALSE, results = "hide", message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE, dev = "svg", echo = TRUE, message = FALSE,
  warning = FALSE, fig.height = 6, fig.width = 1.777777 * 6)
```

```{r, load-library}
library("vegan")
library("readxl")
library("dplyr")
```

We'll need **readxl**, and its `read_xlsx()` function, to read from the 
XLS format files that the example data come as.

The data set itself is quite simple and small, consisting of counts on 23 species from 16 plots, and arise from a randomised complete block designed experiment described by &Scaron;pa&#269;kov&aacute; and colleagues [-@Spackova1998-ad] and analysed by [@Smilauer2014-ac] in their recent book using Canoco v5.

The experiment tested the effects on seedling recruitment to a range of treatments

 * control
 * removal of litter
 * removal of the dominant species *Nardus stricta*
 * removal of litter and moss (mos couldn't be removed without also removing litter)

The treatments were replicated replicated in four, randomised complete blocks.

The data are available from the accompanying website to the book *Multivariate Analysis of Ecological Data using CANOCO 5* [@Smilauer2014-ac]. They are supplied as XLS format files in a ZIP archive. We can read these into R directly from the web with a little bit of effort

```{r, load-data}
## Download the excel data
furl <- "https://bit.ly/seedlings-example"
td <- tempdir()
tf <- tempfile(tmpdir = td, fileext = ".xlsx")
download.file(furl, tf)

## list the sheets in the workbook
excel_sheets(tf)

## read the xlsx file, sheet 2 contains species data, sheet 3 the env
spp <- read_xlsx(tf, sheet = "seedlspe", skip = 1) %>%
  rename("sample_id" = "...1") %>%
  tibble::column_to_rownames("sample_id")
env <- read_xlsx(tf, sheet = "seedldesign")%>%
  rename("sample_id" = "...1") %>%
  tibble::column_to_rownames("sample_id")
```

The `block` variable is currently coded as an integer and needs 
converting to a factor if we are to use it correctly in the analysis

```{r, transform-block}
env <- mutate(env, block = factor(block))
```

The gradient lengths are short,

```{r, decorana}
decorana(spp)
```

motivating the use of redundancy analysis (RDA). Additionally, we may 
be interested in how the raw abundance of seedlings change following 
experimental manipulation, so we may wish to focus on the proportional 
differences between treatments. The first case is handled naturaly by 
RDA. The second case will require some form of standardisation by 
samples, say by sample totals.

First, let's test the first null hypothesis; that there is no effect of 
the treatment on seedling recruitment. This is a simple RDA. We should 
take into account the `block` factor when we assess this model for 
significance. How we do this illustrates two potential approaches to 
performing permutation tests

 1. **design**-based permutations, where how the samples are permuted 
 follows the experimental design, or
 
 2. **model**-based permutations, where the experimental design is 
 included in the analysis directly and residuals are permuted by simple 
 randomisation.

There is an important difference between the two approach, one which 
I'll touch on shortly.

We'll proceed by fitting the model, conditioning on `block` to remove 
between block differences

```{r, fit-conditioned-rda}
mod1 <- rda(spp ~ treatment + Condition(block), data = env)
mod1
```

There is a strong single, linear gradient in the data as evidenced by 
the relative magnitudes of the eigenvalues (here expressed as 
proportions of the total variance)

```{r, eigenvals}
summary(eigenvals(mod1))
```

## Design-based permutations

A *design*-based permutation test of these data would be one conditioned on the `block` variable, by restricting samples to be permuted only *within* the levels of `block`. In this situation, samples are never permuted between blocks, only within. We can set up this type of permutation design as follows

```{r, how1}
h <- with(env, how(blocks = block, nperm = 999))
```

Note that we could use the `plots` argument instead of `blocks` to 
restrict the permutations in the same way, but using `blocks` is 
simpler. I also set the required number of permutations for the test 
here.

`permutations` can take a number of different types of instruction

 1. an object of class `"how"`, whch contains details of a restricted 
 permutation design that `shuffleSet()` from the **permute** package will use to generate permutations from, or
 
 2. a number indicating the number of permutations required, in which 
 case these are simple randomisations with no restriction, unless the 
 `strata` argument is used, or
 
 3. a matrix of user-specified permutations, 1 row per permutation.

To perform the design-based permutation we'll pass `h`, created 
earlier, to `anova()`

```{r, anova1}
set.seed(42)
p1 <- anova(mod1, permutations = h, parallel = 3)
p1
```

Note that I've run this on three cores in parallel; I have four cores on 
my laptop but left one free for the other software I have running.

The overall permutation test indicates no significant effect of 
treatment on the abundance of seedlings. We can test individual axes by 
adding `by = "axis"` to the `anova()` call

```{r, anova1-by-axis}
set.seed(24)
p1axis <- anova(mod1, permutations = h, parallel = 3, by = "axis")
p1axis
```

This confirms the earlier impression that there is a single, linear 
gradient in the data set. A biplot shows that this axis of variation is 
associated with the Moss (& Litter) removal treatment. The variation 
between the other treatments lies primarily along axis two and is 
substantially less than that associated with the Moss & Litter removal.

```{r, biplot, fig = TRUE, fig.cap = "Figure 1: RDA biplot showing species scores and treatment centroids."}
plot(mod1, display = c("species", "cn"), scaling = "sites", type = "n",
     xlim = c(-10.5, 1.5))
text(mod1, display = "species", scaling = 1, cex = 0.8)
text(mod1, display = "cn", scaling = 1, col = "blue", cex = 1.2,
     labels = c("Control", "Litter+Moss", "Litter", "Removal"))
```

In the above figure, I used `scaling = "sites:`, so-called *inter-sample 
distance scaling*, as this best represents the centroid scores, which 
are computed as the treatment-wise average of the sample scores.

## Model-based permutation

The alternative permutation approach, known as model-based 
permutations, and would employ free permutation of residuals after the 
effects of the covariables have been accounted for. This is justified 
because under the null hypothesis, the residuals are freely 
exchangeable once the effects of the covariables are removed. There is 
a clear advantage of model-based permutations over design-based 
permutations; where the sample size is small, as it is here, there 
tends to be few blocks and the resulting design-based permutation test 
relatively weak compared to the model-based version.

It is simple to switch to model-based permutations, be setting the 
blocks indicator in the permutation design to `NULL`, removing the 
blocking structure from the design

```{r, remove-blocks}
setBlocks(h) <- NULL                    # remove blocking
getBlocks(h)                            # confirm
```

Next we repeat the permutation test using the modified `h`

```{r, model-anova1}
set.seed(51)
p2 <- anova(mod1, permutations = h, parallel = 3)
p2
```

The estimated *p* value is slightly smaller now. The difference between 
treatments is predominantly in the Moss & Litter removal with 
differences between the control and the other treatments lying along 
the insignificant axes

```{r, model-anova-by-axis}
set.seed(83)
p2axis <- anova(mod1, permutations = h, parallel = 3, by = "axis")
p2axis
```

## Chages in relative seedling composition

As mentioned earlier, interest is also, perhaps predominantly, in whether any of the treatments have different species composition. To test this hypothesis we standardise by the sample (row) norm using `decostand()`. Alternatively we could have used `method = "total"` to work with proportional abundances. We then repeat the earlier steps, this time using only model-based permutations owing to their greater power.

```{r, standardised-fits}
spp.norm <- decostand(spp, method = "normalize", MARGIN = 1)

mod2 <- rda(spp.norm ~ treatment + Condition(block), data = env)
mod2
eigenvals(mod2) / mod2$tot.chi

set.seed(76)
anova(mod2, permutations = h, parallel = 3)
```

The results suggest no difference in species composition under the experimental manipulation.

## References
