# Birds Data

This data set is on bird assemblages in a montane forest in the Velka Fatra Mountains (Slovak Republic).

```{r, message=FALSE}
library("readr")
library("tidyr")
library("dplyr")
library("tibble")
```

We can load the bird data and process it into two data sets

1. the species abundances, and
2. the associated environmental variables.


```{r}
birds_url <- "https://bit.ly/maed-birds"

birds <- read_csv(birds_url, na = "", skip = 1)

birds_spp <- birds %>%
    mutate(across(AegiCaud:OriOri, ~ replace_na(.x, 0))) %>%
    column_to_rownames("...1")
birds_spp <- birds_spp[, 1:36]

birds_env <- birds %>%
    select(c("...1", Altit:Expos)) %>%
    mutate(Rocks = factor(Rocks), Expos = factor(Expos),
    across(Forest:Slope, ~ ordered(.x))) %>%
    column_to_rownames("...1")
```

The variables are:

* Altitude
* Forest cover
* Forest density
* Amount of broad leaf trees
* E2 layer (scrub layer cover)
* E2 layer conifers
* E1 layer (herb layer cover)
* E1 height
* Slope
* Presence of rocks
* Exposure

The aim of the analysis is to fit a constrained ordination to these data, to identify which of the predictor variables seem to have the best ability to explain variation in bird data.
