---
title: PERMANOVA
author: Gavin Simpson
date: May 31, 2024
output: html_document
editor_options: 
  chunk_output_type: console
---

# Effect of ploughing time on weeds

Post (1986) carried out an experiment to investigate the effect of time of ploughing on the subsequent weed composition in summer barley fields. Three different ploughing times were used. There are 13 weed species observed in 12 plots in a randomized block experiment of four complete blocks of three plots each.

Traditionally, a multivariate ANOVA could have been used to analyses data like this. However, as there are more species (response variables; 13) than there are samples (12), MANOVA cannot be used. Instead, RDA is a useful option.

```{r plough-load}
library("vegan")
library("dplyr")
library("readr")

plough <- read_csv(url("https://bit.ly/plough-spp"), skip = 1) %>%
    rename(label = "...1") %>%
    select(-`...2`) %>%
    janitor::clean_names() %>%
    filter(!is.na(label)) %>%
    tibble::column_to_rownames("label")

plough_design <- read_csv(url("https://bit.ly/plough-design"), skip = 1) %>%
    rename(label = "...1", ploughing_time = "Treatmnt") %>%
    select(-`...2`) %>%
    janitor::clean_names() %>%
    filter(!is.na(label)) %>%
    tibble::column_to_rownames("label")
```

We are interested in the effect of ploughing time and wish to remove any possible effect of the blocks. We cannot achieve this by using `block` as a covariate (in `Condition()`) when we estimate the model using PERMANOVA. Instead we will have to include `block` in our model as a variable. Some zero abundances are recorded as `NA` values, so we must replace those with 0s before we proceed. Instead of log-transforming the species counts, we might use a non-Euclidean dissimilarity coefficient with PERMANOVA instead of RDA.

```{r}
plough <- plough %>%
    mutate(across(che_alb:ape_spi, .fns = tidyr::replace_na, replace = 0))
```

A correct permutation test for these data should keep samples contained at the block level and only permute within, not between, blocks.

```{r}
h <- with(plough_design,
    how(blocks = block, nperm = 999))
```

The model and permutation test are specified as

```{r}
ord <- adonis2(plough ~ ploughing_time, data = plough_design,
    method = "bray", add = "lingoes", permutations = h)
ord
```

Notice how we get a warning/message from *permute*

> Set of permutations < 'minperm'. Generating entire set.

This means that there are not a large number of permutations of these data owing to the block restriction.

Note how with `adonis2()` we don't have the option of using `Condition()` in the model formula. As such we must restrict the permutations to within the blocks of the experimental design. We could also include the `block` effect in the model

```{r}
ord2 <- adonis2(plough ~ ploughing_time + block, data = plough_design,
    method = "bray", add = "lingoes", permutations = h, by = "margin")
ord2
```

We tested the marginal effects of the model terms as the default is to do sequential tests, but we don't want the order in which we added the model terms to the formula to influence the results.

To have greater confidence in the results, we would like to know if there are different variances of the communities in the three groups defined by the ploughing time.

```{r}
disp <- betadisper(vegdist(plough), plough_design$ploughing_time)
disp
```

The `anova()` method performs a classic ANOVA-based test of the dispersions (H0: groups have equal dispersions)

```{r}
anova(disp)
```

Are there any differences between groups in terms of their dispersions?

A permutation test can also be used, instead of assuming the data meet the assumptions of the ANOVA

```{r}
permutest(disp, pairwise = TRUE, permutations = h)
```

Here we asked for pairwise comparisons to investigate whether there are pairs of groups that have different dispersions instead of just looking at the overall test.

Tukey's honest significant differences can also be used to pairwise compare the groups; while this is a parametric test, the confidence intervals are constructed to account for multiple comparisons

```{r}
## Tukey's Honest Significant Differences
(disp_HSD <- TukeyHSD(disp))
plot(disp_HSD)
```

If we want to visualise the estimated dispersions, we can plot the betadisper object

```{r}
plot(disp)
```

An alternative plot is produced using 1 SD data ellipses

```{r}
## with data ellipses instead of hulls
plot(disp, ellipse = TRUE, hull = FALSE) # 1 sd data ellipse
```
