---
title: "Constrained ordination"
author: Gavin L. Simpson
date: September 21, 2022
fontsize: 10pt
classoption: "compress, aspectratio=169"
bibliography: "resources/vegan-refs.bib"
output:
  xaringan::moon_reader:
    css: ['default', 'https://stackpath.bootstrapcdn.com/bootstrap/4.3.1/css/bootstrap.min.css', 'slides.css']
    lib_dir: libs
    nature:
      titleSlideClass: ['inverse','middle','left',my-title-slide]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "macros.js"
      ratio: '16:9'
---
class: inverse middle center big-subsection

```{r setup-options, echo = FALSE, results = "hide", message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE, dev = 'svg', echo = TRUE, message = FALSE, warning = FALSE,
                      fig.height=6, fig.width = 1.777777*6)
library("vegan")
library("ggplot2")
data(varespec)
data(varechem)

## plot defaults
theme_set(theme_minimal(base_size = 16, base_family = 'Fira Sans'))
```

# Welcome

---

# Today's topics

* Constrained ordination
    * Canonical Correspondence Analysis
	* Redundancy Analysis
	* Partial constrained ordination
* Model Building
    * Model selection
* Permutation tests

---
class: inverse middle center subsection

# Constrained Ordination

---
class: inverse middle center big-subsection

# CCA

---

# Canonical Correspondence Analysis

CCA is the constrained form of CA; fitted using `cca()`

Two interfaces for specifying models

* basic; `cca1 <- cca(X = varespec, Y = varechem)`
* formula; `cca1 <- cca(varespec ~ ., data = varechem)`

RDA is the constrained form of PCA; fitted using `rda()`

Formula interface is the more powerful &mdash; *recommended*

---

# Canonical Correspondence Analysis

```{r cca-model}
cca1 <- cca(varespec ~ ., data = varechem)
cca1
```

---

# Redundancy Analysis

```{r rda-model}
rda1 <- rda(varespec ~ ., data = varechem)
rda1
```

---

# The `cca.object`

 * Objects of class `"cca"` are complex with many components
 * Entire class described in `?cca.object`
 * Depending on what analysis performed some components may be `NULL`
 * Used for (C)CA, PCA, RDA, CAP (`capscale()`), and dbRDA (`dbrda()`)

---

# The `cca.object`

`cca1` has a large number of components

* **`$call`** how the function was called
* **`$grand.total`** in (C)CA sum of `rowsum`
* **`$rowsum`** the row sums
* **`$colsum`** the column sums
* **`$tot.chi`** total inertia, sum of Eigenvalues
* **`$pCCA`** Conditioned (partial-ed out) components
* **`$CCA`** Constrained components
* **`$CA`** Unconstrained components
* **`$method`** Ordination method used
* **`$inertia`** Description of what inertia is

---

# The `cca.object`

Depending on how one called `cca()` etc some of these components will be `NULL`

`$pCCA` is only filled in if a *partial* constrained ordination fitted

`rda()` returns objects with classes `"rda"` and `"cca"`, but in most cases those objects work like those of class `"cca"`

The Eigenvalues and axis scores are now spread about the `$CA` and `$CCA` components (also `$pCCA` if a *partial* CCA)

Thankfully we can use *extractor* functions to get at such things

---

# Eigenvalues

Use `eigenvals()` to extract Eigenvalues from a fitted ordination object

```{r eigenvals}
eigenvals(cca1)
```

---

# Example

- Fit a CCA model to the lichen pasture data. The model should include, N, P, and K only.
- Save the model in object `mycca1`
- How much variance is explained by this model?
- Extract the eigenvalues, how many constrained axes are there?

---

```{r your-turn-fit-cca-1}
library("vegan")
data(varechem, varespec)
```

--

```{r your-turn-fit-cca-2}
mycca1 <- cca(varespec ~ N + P + K, data = varechem)
mycca1
```

---

```{r your-turn-fit-cca-3}
ev <- eigenvals(mycca1, model = "constrained")
head(ev)
length(ev)
```


---

# Extracting axis scores

To extract a range of scores from  a fitted ordination use `scores()`

* takes an ordination object as the first argument
* `choices` &mdash; which axes? Defaults to `c(1,2)`
* `display` &mdash; which type(s) of scores to return
    - `"sites"` or `"wa"`: scores for samples in response matrix
    - `"species"`: scores for variables/columns in response
    - `"lc"`: linear combination site scores
    - `"bp"`: biplot scores (coords of arrow tip)
    - `"cn"`: centroid scores (coords of factor centroids)

---

# Extracting axis scores

```{r scores}
str(scores(cca1, choices = 1:4, display = c("species","sites")), max = 1)
head(scores(cca1, choices = 1:2, display = "sites"))
```

---

# Scalings&hellip;

When we draw the results of many ordinations we display 2 or more sets of data

Can't display all of these and maintain relationships between the scores

*Solution* scale one set of scores relative to the other via the `scaling` argument

---

# Scalings&hellip;

* `scaling = 1` &mdash; Focus on sites, scale site scores by $\lambda_i$
* `scaling = 2` &mdash; Focus on species, scale species scores by $\lambda_i$
* `scaling = 3` &mdash; Symmetric scaling, scale both scores by $\sqrt{\lambda_i}$
* `scaling = -1` &mdash; As above, but
* `scaling = -2` &mdash; For `cca()` multiply results by $\sqrt{(1/(1-\lambda_i))}$
* `scaling = -3` &mdash; this is Hill's scaling
* `scaling < 0` &mdash; For `rda()` divide species scores by species' $\sigma$
* `scaling = 0` &mdash; raw scores

```{r scaling-example, results = "hide"}
scores(cca1, choices = 1:2, display = "species", scaling = 3)
```

---

# Scalings&hellip;

Thankfully we can use alternative descrpitors to extract scores:

- `"none"`
- `"sites"`
- `"species"`
- `"symmetric"`

Two modifiers select negative scores depending on whether the model is CCA or RDA:

- `hill = TRUE`
- `correlation = TRUE`

---

# Example

- Using the CCA model you fitted, extract the site scores for axes 2 and 3 with Hill's scaling, focusing on the sites

--

```{r your-turn-cca-4}
scrs <- scores(mycca1, display = "sites", choices = c(2,3),
               scaling = "sites", hill = TRUE)
head(scrs)
```

---

# Partial constrained ordinations

*Partial* constrained ordinations remove the effect of one or more variables *then* fit model of interest

Argument `Z` is used for a data frame of variables to partial out

```{r partial-ordination-1}
pcca <- cca(X = varespec,
            Y = varechem[, "Ca", drop = FALSE],
            Z = varechem[, "pH", drop = FALSE])
```
Or with the formula interface use the `Condition()` function

```{r partial-ordination-2}
pcca <- cca(varespec ~ Ca + Condition(pH), data = varechem) ## easier!
```

---

# Partial constrained ordinations

```{r partial-ordination-3}
pcca <- cca(varespec ~ Ca + Condition(pH), data = varechem) ## easier!
pcca
```

---

# Triplots

.row[
.col-6[

Triplots will generally produce a mess; we can really only display a couple of bits approximately anyway Trying to cram three things in is a recipe for a mess&hellip; but we can do it
]
.col-6[
```{r triplot-1, fig.height = 6, fig.width = 6}
plot(cca1)
```
]
]

---
class: inverse middle center subsection

# Model building

---

# Building constrained ordination models

If we don't want to think it's easy to fit a poor model with many constraints &mdash; That's what I did with `cca1` and `rda1`

Remember, CCA and RDA are *just regression methods* &mdash; everything you know about regression applies here

A better approach is to *think* about the important variables and include only those

The formula interface allows you to create interaction or quadratic terms easily (though be careful with latter)

It also handles factor or class constraints automatically unlike the basic interface

---

# Building constrained ordination models

```{r cca-model-build1}
vare.cca <- cca(varespec ~ Al + P*(K + Baresoil), data = varechem)
vare.cca
```

---

# Building constrained ordination models

For CCA, RDA etc we have little choice but to do

 1. Fit well-chosen set of candidate models & compare, or
 2. Fit a *full* model of well-chosen variables & then do stepwise selection

Automatic approaches to model building should be used cautiously!

The standard `step()` function can be used as *vegan* provides two helper methods, `deviance()` and `extractAIC()`, used by `step()`

*vegan* also provides methods for class `"cca"` for `add1()` and `drop1()`

---

# Variance inflation factors

*Linear* dependencies between constraints can be investigated with VIF

VIF is a measure of how much the variance of $\hat{\beta}_j$ is inflated by presence of other covariates

Lots of rules of thumb

 * VIF >= 20 indicates *strong collinearity* in constraints
 * VIF >= 10 potentially of concern & should be looked at

Computed via `vif.cca()`

```{r vif-cca1}
vif.cca(cca1)
```

---

# Stepwise selection in CCA

`step()` uses AIC which is a fudge for RDA/CCA &mdash; use `ordistep()`

1. Define an upper and lower model scope, say the full model and the null model
2. To step from the lower scope or null model we use

```{r stepwise-1}
upr <- cca(varespec ~ ., data = varechem)
lwr <- cca(varespec ~ 1, data = varechem)
set.seed(1)
mods <- ordistep(lwr, scope = formula(upr), trace = 0)
```

`trace = 0` is used here to turn off printing of progress

Permutation tests are used (more on these later); the theory for an AIC for ordination is somewhat loose

---

# Stepwise selection in CCA

The object returned by `step()` is a standard `"cca"` object with an extra component `$anova`

```{r stepwise-cca}
mods
```

---

# Stepwise selection in CCA

The `$anova` component contains a summary of the steps involved in automatic model building

```{r stepwise-anova}
mods$anova
```

1. `Al` added first, then
2. `P`, followed by
3. `K`, then stopped

---

# Stepwise selection in CCA

Step-wise model selection is fairly fragile; if we start from the full model we won't end up with the same final model

```{r stepwise-reverse, results = "hide"}
mods2 <- step(upr, scope = list(lower = formula(lwr), upper = formula(upr)), trace = 0,
              test = "perm")
mods2
```

---

# Stepwise selection in CCA

```{r stepwise-reverse}
```

---

# Adjusted $R^2$ for ordination models

Ordinary $R^2$ is biased for the same reasons as for a linear regression

 * adding a variable to constraints will increase $R^2$
 * the larger the number of constraints in the model the larger $R^2$, is due to random correlations

Can attempt to account for this bias via an *adjusted* $R^2$ measure

---

# Adjusted $R^2$ for ordination models

Can attempt to account for this bias via an *adjusted* $R^2$ measure

$$R^2_{adj} = 1 - \frac{n - 1}{n - m - 1}(1 - R^2)$$

* $n$ is number of samples $m$ is number of constraints (model degrees of freedom)
* Can be used up to $\sim M > n/2$ before becomes too conservative
* Can be negative

```{r rsq-cca1}
RsquareAdj(cca1)
```

---

# Stepwise selection via adjusted $R^2$

Problems with stepwise selection are myriad. Affects RDA, CCA, etc

Blanchet *et al* (2008) proposed a two-step solution for models where $R^2_{adj}$ makes sense

---

# Stepwise selection via adjusted $R^2$

* *Global test* of all constraints
    * Proceed **only** if this test is significant
    * Helps prevent inflation of overall type I error
* Proceed with forward selection, but with *two* stopping rules
    * Usual significance threshold $\alpha$
    * The global $R^2_{adj}$
    * Stop if next candidate model is non-significant or if $R^2_{adj}$ exceeds the global $R^2_{adj}$

Available in `ordiR2step()`

---

# Stepwise selection via adjusted $R^2$

```{r stopping-rules}
ordiR2step(lwr, upr, trace = FALSE)
```

---
class: inverse middle center subsection

# Permutation tests

---

# Permutation tests in vegan

RDA has lots of theory behind it, CCA bit less. However, ecological/environmental data invariably violate what little theory we have

Instead we use permutation tests to assess the *importance* of fitted models &mdash; the data are shuffled in some way and the model refitted to derive a Null distribution under some hypothesis of *no effect*

---

# Permutation tests in vegan

What *is* shuffled and *how* is of **paramount** importance for the test to be valid

* No conditioning (partial) variables then rows of the species data are permuted
* With conditioning variables, two options are available, both of which *permute residuals* from model fits
    * The *full model* uses residuals from model $Y = X + Z + \varepsilon$
    * The *reduced model* uses residuals from model $Y = Z + \varepsilon$
* In **vegan** which is used can be set via argument `model` with `"direct"`, `"full"`, and `"reduced"` respectively

---

# Permutation tests in vegan

A test statistic is required, computed for observed model & each permuted model

**vegan** uses a pseudo $F$ statistic

$$F=\frac{\chi^2_{model} / df_{model}}{\chi^2_{resid} / df_{resid}}$$

Evaluate whether $F$ is unusually large relative to the null (permutation) distribution of $F$

---

# Permutation tests in vegan

.row[
.col-6[
```{r permustats-1, results = "hide"}
pstat <- permustats(anova(cca1))
summary(pstat)
```
.small[
```{r permustats-1, echo = FALSE}
```
]
]
.col-6[
```{r permustats-2, fig.width = 6, fig.height = 6}
densityplot(pstat)
```
]
]

???

The summary method of permustats estimates the standardized effect sizes (SES) as the difference
of observed statistic and mean of permutations divided by the standard deviation of permutations
(also known as z-values). It also prints the the mean, median, and limits which contain interval
percent of permuted values. With the default (interval = 0.95), for two-sided test these are (2.5%,
97.5%) and for one-sided tests either 5% or 95% quantile and the p-value depending on the test
direction. The mean, quantiles and z values are evaluated from permuted values without observed
statistic, but the p-value is evaluated with the observed statistic. 

---

# Permutation tests in vegan: `anova()`

* The main user function is the `anova()` method
* It is an interface to the lower-level function `permutest.cca()`
* At its most simplest, the `anova()` method tests whether the **model** as a whole is significant

---

# Permutation tests in vegan: `anova()`

$$F = \frac{1.4415 / 14}{0.6417 / 9} = 1.4441$$

```{r cca-anova}
set.seed(42)
(perm <- anova(cca1))
```

---

# Permutation tests in vegan: `anova()`

`anova.cca()` has a number of arguments

```{r anova-args}
args(anova.cca)
```

`object` is the fitted ordination

`permutations` controls what is permuted and how

`by` determines what is tested; the default is to test the model

---

# Types of permutation test in vegan

A number of types of test can be envisaged

 * Testing the overall significance of the model
 * Testing constrained (canonical) axes
 * Testing individual model terms *sequentially*
 * The *marginal* effect of a single variable

The first is the default in `anova()`

The other three can be selected via the argument `by`

---

# Testing canonical axes

* The constrained (canonical) axes can be individually tests by specifying `by = "axis"`
* The first axis is tested in terms of variance explained compared to residual variance
* The second axis is tested after partialling out the first axis&hellip;
* and so on

---

# Testing canonical axes

```{r anova-by-axis}
set.seed(1)
anova(mods, by = "axis")
```

---

# Testing terms sequentially

* The individual terms in the model can be tested using `by = "terms"`
* The terms are assessed in the order they were specified in the model, sequentially from first to last
* Test is of the additional variance explained by adding the $k$th variable to the model
* **Ordering of the terms** will affect the results

---

# Testing terms sequentially

```{r anova-by-term}
set.seed(5)
anova(mods, by = "terms")
```

---

# Testing terms marginal effects

* The marginal *effect* of a model term can be assessed using `by = "margin"`
* The marginal *effect* is the effect of a particular term when all other model terms are included in the model

---

# Testing terms marginal effects

```{r anova-by-margin}
set.seed(10)
anova(mods, by = "margin")
```

---

# Spring meadow vegetation

Example & data taken from Leps & Smilauer (2014), Case Study 2

Spring fen meadow vegetation in westernmost Carpathian mountains

```{r meadows-setup}
## load vegan
library("vegan")

## load the data
spp <- read.csv("data/meadow-spp.csv", header = TRUE, row.names = 1)
env <- read.csv("data/meadow-env.csv", header = TRUE, row.names = 1)
```

---

# Spring meadow vegetation

CCA a reasonable starting point as the gradient is long here (check with `decorana()` if you want)

```{r meadows-cca-full}
m1 <- cca(spp ~ ., data = env)
set.seed(32)
anova(m1)
```

---

# Spring meadow vegetation

.row[
.col-6[
```{r meadows-cca-full-triplot, fig.show = "hide"}
plot(m1)
```
]
.col-6[

```{r meadows-cca-full-triplot, fig.height = 6, fig.width = 6, echo = FALSE}
```
]
]

---

# Spring meadow vegetation

```{r meadows-cca-stepwise}
set.seed(67)
lwr <- cca(spp ~ 1, data = env)
( m2 <- ordistep(lwr, scope = formula(m1), trace = FALSE) )
```

---

# Spring meadow vegetation

.row[
.col-6[
```{r meadows-cca-reduced-triplot, fig.show = "hide"}
plot(m2)
```
]
.col-6[

```{r meadows-cca-reduced-triplot, fig.height = 6, fig.width = 6, echo = FALSE}
```
]
]

---

# Spring meadow vegetation

```{r meadows-cca-anova}
m2$anova
```

---

# Spring meadow vegetation

Alternative is RDA with a transformation

.row[
.col-6[
```{r meadows-rda}
spph <- decostand(spp, method = "hellinger")
m3 <- rda(spph ~ ., data = env)
lwr <- rda(spph ~ 1, data = env)
m4 <- ordistep(lwr, scope = formula(m3),
               trace = FALSE)
```
]
.col-6[
.small[
```{r meadows-rda-print}
m4
```
]
]
]
---

# Spring meadow vegetation

.row[
.col-6[
```{r meadows-rda-reduced-triplot, fig.show = "hide"}
plot(m4)
```
]
.col-6[

```{r meadows-rda-reduced-triplot, fig.height = 6, fig.width = 6, echo = FALSE}
```
]
]

---

# Spring meadow vegetation

Stepwise using $R^2_{adj}$

```{r meadows-rda-adjrsquare}
m5 <- ordiR2step(lwr, scope = formula(m3), trace = FALSE)
m5$anova
```

---

# References

.small[
* Anderson, M.J., 2001. A new method for non-parametric multivariate analysis of variance. Austral Ecol. 26, 32&ndash;46
* Anderson, M.J., 2006. Distance-based tests for homogeneity of multivariate dispersions. Biometrics 62, 245&ndash;253
* Anderson, M.J., Walsh, D.C.I., 2013. PERMANOVA, ANOSIM, and the Mantel test in the face of heterogeneous dispersions: What null hypothesis are you testing? Ecol. Monogr. 83, 557&ndash;574
* Anderson, M.J., Walsh, D.C.I., Robert Clarke, K., Gorley, R.N., Guerra-Castro, E., 2017. Some solutions to the multivariate Behrens-Fisher problem for dissimilarity-based analyses. Aust. N. Z. J. Stat. 59, 57&ndash;79
* Blanchet, F.G., Legendre, P., Borcard, D., 2008. Forward selection of explanatory variables. Ecology 89, 2623&ndash;2632
* Legendre, P., Anderson, M.J., 1999. Distance-based redundancy analysis: testing multispecies responses in multifactorial ecological experiments. Ecol. Monogr. 69, 1&ndash;24
* McArdle, B.H., Anderson, M.J., 2001. Fitting Multivariate Models to Community Data: A Comment on Distance-Based Redundancy Analysis. Ecology 82, 290&ndash;297
* Warton, D.I., Wright, S.T., Wang, Y., 2012. Distance-based multivariate analyses confound location and dispersion effects. Methods Ecol. Evol. 3, 89&ndash;101
]
