---
title: Principal response curves
author: Gavin Simpson
date: November 24, 2022
output: html_document
---

```{r ohraz}
library("vegan")
library("readr")
library("dplyr")
library("forcats")

spp <- read_csv(url("https://bit.ly/ohraz-spp")) %>%
    rename(label = "...1") %>%
    janitor::clean_names()

molinia <- spp %>%
    select(label:molicaer)

spp <- spp %>%
    select(-molicaer) %>%
    tibble::column_to_rownames("label")

env <- read_csv(url("https://bit.ly/ohraz-env")) %>%
    rename(label = "...1") %>%
    mutate(across(c(mowing:removal, plotid), ~ factor(.x))) %>%
    tibble::column_to_rownames("label")
```

We need to recode the variables to make this easier to understand for PRC

```{r}
env2 <- env %>%
mutate(mowed = fct_recode(mowing, mowed = "Yes", unmowed = "No"),
removed = fct_recode(removal, control = "No", removed = "Yes"),
fertilized = fct_recode(fertilizer, fertilized = "Yes", unfertilized = "No"),
removed = fct_relevel(removed, "control"),
treatment = fct_cross(removed, mowed, fertilized),
fyear = factor(year))
```

And we should log transform the species data as this is an RDA:
```{r}
log_spp <- log1p(spp)
```

The PRC can then be fitted using

```{r}
ohraz_prc <- prc(log_spp, env2$treatment, env2$fyear)
```

The permutation deisgn to use is

```{r}
h <- with(env2, how(plots = Plots(strata = plotid), nperm = 999))
```

And with this we can now test the first PRC (RDA) axis

```{r}
anova(ohraz_prc, permutations = h, first = TRUE)
```

**Are the time structured effects of the treatment significant?**

We might also restrict the permutations within plots for time series

```{r}
h_time <- with(env2, how(plots = Plots(strata = plotid),
    within = Within(type = "series"), nperm = 999))
```

This gives a stricter test, especially if there is residual autocorrelation

```{r}
anova(ohraz_prc, permutations = h_time, first = TRUE)
```

Does the effect remain after performing the stricter test?

Plot the PRC, without the species scores

```{r}
plot(ohraz_prc, species = FALSE, legpos = "bottomleft", col = 1:8, lty = 1, lwd = 2)
```

**Which treatment levels/combinations lead to the largest changes in the species composition by the end of the experiment?**

We can plot the species scores on a separate plot, retaining only those species that have a fit greater than 5%

```{r}
fit <- goodness(ohraz_prc, display = "species", choices = 1, scaling = "symmetric")
scrs <- scores(ohraz_prc, display = "species", choices = 1, scaling = "symmetric")
linestack(scrs[fit > 0.05, , drop = FALSE])
axis(side = 2)
```

Suggest which species were affected most in the treatments that affected the composition the most.