---
title: "Other topics"
author: Gavin L. Simpson
date: November 25, 2022
fontsize: 10pt
classoption: "compress, aspectratio=169"
bibliography: "resources/vegan-refs.bib"
output:
  xaringan::moon_reader:
    css: ['default', 'https://stackpath.bootstrapcdn.com/bootstrap/4.3.1/css/bootstrap.min.css', 'slides.css']
    lib_dir: libs
    nature:
      titleSlideClass: ['inverse','middle','left',my-title-slide]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "macros.js"
      ratio: '16:9'
---
class: inverse middle center big-subsection

```{r setup-options, echo = FALSE, results = "hide", message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE, dev = 'svg', echo = TRUE, message = FALSE, warning = FALSE,
                      fig.height=6, fig.width = 1.777777*6)
library("vegan")
library("ggplot2")
library("dplyr")
library("tibble")
library("janitor")
library("readr")
data(varespec)
data(varechem)

## plot defaults
theme_set(theme_minimal(base_size = 16, base_family = 'Fira Sans'))
```

# Welcome

---

# Today's topics

* PERMANOVA
* Distance-based RDA
* Co-correspondence analysis
* Diagnostics

---
class: inverse middle center big-subsection

# PERMANOVA

---

# MANOVA

MANOVA is the multivariate form of ANOVA

* Multivariate response data
* Categorical predictor variables

Decompose variation in the responses into

1. variation within groups
2. variation between groups

Test to see if two is unusually large relative to H<sub>0</sub>

---

# PERMANOVA

Doing that test requires lots of assumptions that rarely hold for ecological data

PERMANOVA: Permutational multivariate analysis of variance

Avoids most of these issues through the use of permutation tests

Directly decomposes a dissimilarity matrix into

1. variation within groups
2. variation between groups

---

# PERMANOVA *sensu stricto*

*vegan* has four different ways to do essentially do this kind of analysis

1. `adonis()` &mdash; implements Anderson (2001)
2. `adonis2()` &mdash; implements McArdle & Anderson (2001)
3. `dbrda()` &mdash; implementation based on McArdle & Anderson (2001)
4. `capscale()` &mdash; implements Legendre & Anderson (1999)

Be careful with `adonis()` as it allows only sequential tests

A difference between the functions is how they treat negative eigenvalues

---

# The PERMANOVA idea

.center[
```{r permanova-idea-plot, echo = FALSE}
data(varespec)
     
## Bray-Curtis distances between samples
dis <- vegdist(varespec)
     
## First 16 sites grazed, remaining 8 sites ungrazed
groups <- factor(c(rep(1,16), rep(2,8)), labels = c("grazed","ungrazed"))
     
## Calculate multivariate dispersions
mod <- betadisper(dis, groups)
plot(mod)
```
]

---

# PERMANOA &mdash; `adonis2()`

```{r adonis2-by-terms}
data(dune, dune.env)
adonis2(dune ~ Management*A1, data = dune.env, by = "terms")
```

---

# PERMANOA &mdash; `adonis2()`

```{r adonis2-by-terms-flipped}
data(dune, dune.env)
adonis2(dune ~ A1*Management, data = dune.env, by = "terms")
```

---

# PERMANOA &mdash; `adonis2()`

```{r adonis2-by-margin}
data(dune, dune.env)
adonis2(dune ~ Management*A1, data = dune.env, by = "margin")
```

--

The interaction is the only term that isn't *marginal* to other terms; not significant

---

# PERMANOA &mdash; `adonis2()`

```{r adonis2-margin-2}
adonis2(dune ~ Management + A1, data = dune.env, by = "margin")
```

---

# The dispersion problem

Anderson (2001) noted that PERMANOVA could confound *location* & *dispersion* effects

If one or more groups are more variable &mdash; dispersed around the centroid &mdash; than the others, this can result in a false detection of a difference of means &mdash; a *location* effect

Same problem affects *t* tests, ANOVA

Warton et al (2012)
Anderson & Walsh (2013)
Anderson *et al* (2017)

---

# Dispersion

.center[
```{r permanova-idea-plot, echo = FALSE}
```
]

---

# Test for dispersion effects

Marti Anderson (2006) developed a test for multivariate dispersions &mdash; PERMDISP2

1. Calculate how far each observation is from its group median (or centroid)
2. Take the absolute values of these distances-to-medians
3. Do an ANOVA on the absolute distances with the *groups* as covariates
4. Test the H<sub>0</sub> of equal absolute distances to median among groups using a permutation test

In *vegan* this is `betadisper()`

---

# Test for dispersion effects

.row[
.col-6[
.not-so-small[
```{r permdisp}
data(varespec)
dis <- vegdist(varespec) # Bray-Curtis distances
## First 16 sites grazed, remaining 8 sites ungrazed
groups <- factor(c(rep(1,16), rep(2,8)),
                 labels = c("grazed","ungrazed"))

mod <- betadisper(dis, groups)
mod
```
]
]
.col-6[
```{r permdisp-plot, fig.height = 6, fig.width = 6}
boxplot(mod)
```
]
]

---

# Test for dispersions

.row[

.col-6[
.smaller[
```{r permdisp-anova}
set.seed(25)
permutest(mod)
```
]
]

.col-6[
```{r permdisp-plot-it, fig.width = 6, fig.height = 6}
plot(mod)
```

]
]

---

# Test for dispersions

```{r permdisp-anova-2}
set.seed(4)
permutest(mod, pairwise = TRUE)
```

---

# Test for locations with non-equal dispersion?

Marti Anderson & colleagues (2017) have proposed a solution that is related to the Berens-Fisher problem

This is in Primer but not yet in *vegan*

<https://github.com/vegandevs/vegan/issues/344>

---
class: inverse middle center big-subsection

# Distance-based RDA

---

# Distance-based RDA

Multiple models that all do something Similar

1. `adonis()` (deprecated)
2. `adonis2()`
3. `capscale()`
4. `dbrda()`

They all do essentially the same thing, but they do it differently & have slightly different behaviour

---

# Distance-based RDA

Distance-based RDA (db-RDA) is a constrained form of principal coordinates analysis (PCO)

It is similar to RDA but allows for non-Euclidean dissimilarity indices

In *vegan*, db-RDA is implemented in `dbrda()`

---

# Constrained analysis of principal coordinates

`capscale()` is *another* constrained form of PCO due to Legendre & Anderson (1999)

It is *very* similar to `dbrda()` but handles the negative eigenvalue problem in a different way &mdash; it ignores them!

---

# Constrained analysis of principal coordinates

`capscale()` works by

1. convert the response data into dissimilarities
2. apply PCO on the dissimilarities, take the PCO sample (site) scores as *new* response data
3. fit `rda()` to the *new* response data and predictor variables as constraints

Essentially, we embed the dissimilarities in a Euclidean space using PCO, and then we use RDA on this highly transformed response data

---

# Distance-based RDA

db-RDA foregoes step 2., and directly decomposes the dissimilarities into components explained by each term in the model

Negative eigenvalues resulting from non-metric dissimilarity coefficients are handled via

1. square-root transform of the dissimilarities, or
2. adding a constant to the dissimilarities using methods `"lingoes"` (default, preferred) or `"cailliez"`

db-RDA is based on the ideas in McArdle & Anderson (2001)

--

Err&hellip; isn't that what `adonis2()` was developed to do?

--

*Yes*, but&hellip;

---

# Distance-based RDA

`adonis2()` was a ground up redevelopment of the `adonis()` implementation and as such it retains many of the arguments and concepts of PERMANOVA, just updated to use the direct decomposition of dissimilarities

`dbrda()` inherits from `rda()` and `cca()` and as a result has expanded set of capability

`dbrda()` can use `Condition()` in the formula to fit partial db-RDA

`Condition()` is often needed to provide correct restricted perumtation tests

---

# Distance-based RDA

The equivalent model to `adonis2()` in `dbrda()`-form is

```{r adonis2-by-margin-as-db-rda}
data(dune, dune.env)
dune_dbrda <- dbrda(dune ~ Management * A1, data = dune.env,
    method = "bray")
```

because they have different default `method` values

---
class: inverse middle center subsection

# Ohraz Case Study

---

# Restricted permutations | Ohraz

Analyse the Ohraz data Case study 5 of Leps & Smilauer

Repeated observations of composition from an experiment

* Factorial design (3 replicates)
* Treatments: fertilisation, mowing, *Molinia* removal

Test 1 of the hypotheses

> There are *no* directional changes in species composition in time that are common to all treatments or specific treatments

---

# Restricted permutations | Ohraz

```{r}
spp <- read_csv(url("https://bit.ly/ohraz-spp")) %>%
    rename(label = "...1") %>%
    janitor::clean_names()

molinia <- spp %>%
    select(label:molicaer)

spp <- spp %>%
    select(-molicaer) %>%
    column_to_rownames("label")

env <- read_csv(url("https://bit.ly/ohraz-env")) %>%
    rename(label = "...1") %>%
    mutate(across(c(mowing:removal, plotid), ~ factor(.x))) %>%
    column_to_rownames("label")
```

---

# Distance-based RDA | Ohraz

```{r worked-example-devel-2-with-dbrda}
ohraz_dbrda <- dbrda(spp ~ year +
    year:mowing + year:fertilizer + year:removal +
    Condition(plotid), data = env, method = "bray", add = "lingoes")
h <- how(within = Within(type = "none"),
    plots = Plots(strata = env$plotid, type = "free"))
set.seed(42)
anova(ohraz_dbrda, permutations = h, model = "reduced")
```

---
class: inverse middle center big-subsection

# Co-CA

---

# Co-correspondence analysis

---
class: inverse middle center big-subsection

# PRC

---

# Principal Response Curves

---
class: inverse middle center big-subsection

# Other stuff

---

# Diagnostics for constrained ordinations

**vegan** provides a series of diagnostics to help assess the model fit

* `goodness()`
* `inertcomp()`
* `spenvcor()`
* `intersetcor()`
* `vif.caa()`

---

# Diagnostics | goodness of fit

`goodness()` computes a goodness of fit statistic for species or sites, controlled by argument `display`

Gives the cumulative proportion of variance explained by each axis

```{r goodness}
upr <- cca(varespec ~ ., data = varechem)
lwr <- cca(varespec ~ 1, data = varechem)
set.seed(1)

mods <- ordistep(lwr, scope = formula(upr), trace = 0)
head(goodness(mods))
```

---

# Diagnostics | inertia decomposition

`inertcomp()` decomposes the variance in samples or species in partial, constrained, and unconstrained components

* `statistic = "explained` (default) gives the decomposition in terms of variance
* `statistic = "distance"` gives decomposition in terms of the the residual distance

```{r inertcomp}
head(inertcomp(mods, proportional = TRUE))
```

---

# Diagnostics | species-environment correlations

`spenvcor()` returns the (weighted) correlation between the weighted average-based and the linear combination-based sets of site scores

A *poor* measure of goodness of fit. Sensitive to

* outliers (like all correlations)
* overfitting (using too many constraints)

Better models can have poorer species-environment correlations

```{r spenvcor}
spenvcor(mods)
```

---

# Diagnostics | interset correlations

`intersetcor()` returns the (weighted) correlation between the weighted average-based site scores and each constraint variable

Another *poor* diagnostic

* correlation based
* focuses on a single constraint--axis combination at a time

```{r intersetcor}
intersetcor(mods)
```

Vector fitting (`envfit()`) or biplot scores (`scores(model, display = "bp")`) are better alternatives

---

# References

.smaller[
* Anderson, M.J., 2001. A new method for non-parametric multivariate analysis of variance. Austral Ecol. 26, 32&ndash;46
* Anderson, M.J., 2006. Distance-based tests for homogeneity of multivariate dispersions. Biometrics 62, 245&ndash;253
* Anderson, M.J., Walsh, D.C.I., 2013. PERMANOVA, ANOSIM, and the Mantel test in the face of heterogeneous dispersions: What null hypothesis are you testing? Ecol. Monogr. 83, 557&ndash;574
* Anderson, M.J., Walsh, D.C.I., Robert Clarke, K., Gorley, R.N., Guerra-Castro, E., 2017. Some solutions to the multivariate Behrens-Fisher problem for dissimilarity-based analyses. Aust. N. Z. J. Stat. 59, 57&ndash;79
* Blanchet, F.G., Legendre, P., Borcard, D., 2008. Forward selection of explanatory variables. Ecology 89, 2623&ndash;2632
* Legendre, P., Anderson, M.J., 1999. Distance-based redundancy analysis: testing multispecies responses in multifactorial ecological experiments. Ecol. Monogr. 69, 1&ndash;24
* McArdle, B.H., Anderson, M.J., 2001. Fitting Multivariate Models to Community Data: A Comment on Distance-Based Redundancy Analysis. Ecology 82, 290&ndash;297
* Warton, D.I., Wright, S.T., Wang, Y., 2012. Distance-based multivariate analyses confound location and dispersion effects. Methods Ecol. Evol. 3, 89&ndash;101
]
