---
title: Vltava forest data
author: Gavin Simpson
date: April 24, 2023
output: html_document
---
# Vltava river valley forest vegetation

Vegetation plots, located at even distances along transects following the steep valley slopes of Vltava river valley, collected during 2001-2003. Each transect starts at the valley bottom and ends up at the upper part of the valley slope. Altogether 97 plots located in 27 transects were collected, each of the size 10x15 m. In each plot, all tree, shrub and herb species were recorded and their abundances were estimated using 9-degree ordinal Braun-Blanquette scale (these values were consequently transformed into percentage scale).

```{r, results = "hide"}
library("vegan")

v_url <- "https://bit.ly/vltava-spp"
vltava <- read.delim (url(v_url), row.names = 1)
head(vltava)
```

We fit 4 clustering methods:

```{r transform}
vltava_sqrt <- sqrt(vltava)
dis <- vegdist(decostand(vltava, method = "hellinger"),
    method = "euclidean")

cl_single   <- hclust (dis, method = 'single')
cl_complete <- hclust (dis, 'complete')
cl_average  <- hclust (dis, "average")
cl_ward     <- hclust (dis, "ward.D2")
```

Plot the dendrograms

```{r}
layout(matrix(1:4, ncol = 2))
plot(cl_single, hang = -0.1)
plot(cl_complete, hang = -0.1)
plot(cl_average, hang = -0.1)
plot(cl_ward, hang = -0.1)
layout(1)
```

Now look at *k*-means

```{r}
set.seed(42)
vltava_hell <- decostand(vltava, method = "hellinger")
cas_km <- cascadeKM(vltava_hell, 1, 10)


set.seed(42)
cas_km <- cascadeKM(sqrt(vltava), 1, 10)
```

We can plot this

```{r}
plot(cas_km)
```

### Cophenetic correlation

The cophenetic distances can be computed from the dendrogram using `cophenetic()`. Then we can compute the cophenetic correlation, the correlation between the original dissimilarities and the cophenetic distances using `cor()`

```{r}
cop <- cophenetic(cl_ward)
cor(cop, dis)
```

## Diversity

Refer back to the slides and compute the following diversity measure

* Richness

* Shannon-Wiener

* Simpson's

* Hills N2